---
title: "PainSense Data Transformation"
author: "Kevin Matnok Hansen and Lovisa Johansson"
format: docx
editor: visual
engine: knitr
---

# Purpose

This document describes the data cleaning and transformation process for the PainSense baseline dataset. Raw survey data exported from Sunet Survey (Excel format) is restructured into a tidy dataset ready for statistical analysis.

The pipeline is built on three principles:

1.  **Classification-driven processing**: variables are grouped into *exclusive*, *non-exclusive*, and *matrix* types, with classification verified against VariableView metadata before any data is touched.
2.  **Traceability**: each derived variable links back to its source variable and questionnaire item.
3.  **Documented scoring**: instrument scoring rules are cited and applied consistently.

------------------------------------------------------------------------

## Chapter 1: Initialization

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/nepet/Documents/Data Management Sahlgrenska Academy")
```

```{r}
library(tidyverse)
library(readxl)
library(rlang)
library(labelled)
library(openxlsx)
```

The Excel file has two sheets: **VariableView** (codebook with variable names, labels, types, and value codes) and **Data** (one row per respondent).

```{r}
filepath <- "data/Testfile.xlsx"
vv        <- read_excel(filepath, sheet = "VariableView")
processed <- read_excel(filepath, sheet = "Data")
```

------------------------------------------------------------------------

## Chapter 2: Variable Classification

Variables with pattern `VARnn_mm` are grouped by base name (`VARnn`) and assigned to one of three classes:

| Class | Description | Processing |
|------------------------|------------------------|------------------------|
| **exclusive** | Single-select (radio button) | Collapsed to one numeric column |
| **non_exclusive** | Multi-select (checkbox) | Kept as binary 0/1 columns |
| **matrix** | Independent sub-items with own scale | Left unchanged |

Classification is based on `PainSensPROM_Baseline_DataManagement.docx` and verified against `Enkät_T1a_Gen_230102.pdf`. To reclassify a variable, move it between the three vectors below.

```{r}
exclusive <- c(
  "VAR04",  # Q1a  juridiskt kön: Man / Kvinna
  "VAR05",  # Q1b  könsidentitet: Man / Kvinna / Annat
  "VAR08",  # Q4   utbildning
  "VAR09",  # Q5   födelseland
  "VAR11",  # Q7a  smärta senaste 7 dagar: Ja / Nej
  "VAR13",  # Q7c  NRS medel 7d (0–10)
  "VAR14",  # Q8   NRS vila 7d (0–10)
  "VAR15",  # Q9   NRS rörelse 7d (0–10)
  "VAR16",  # Q10  smärtfrekvens (5 alternativ)
  "VAR17",  # Q11  smärtduration (5 alternativ)
  "VAR18",  # Q12  periodisk / ihållande (2 alternativ)
  "VAR25",  # Q19a HAD – ångest
  "VAR26",  # Q19b HAD – depression
  "VAR27",  # Q19c HAD – ångest
  "VAR28",  # Q19d HAD – depression
  "VAR29",  # Q19e HAD – ångest
  "VAR30",  # Q19f HAD – depression
  "VAR31",  # Q19g HAD – ångest
  "VAR32",  # Q19h HAD – depression
  "VAR33",  # Q19i HAD – ångest
  "VAR34",  # Q19j HAD – depression
  "VAR35",  # Q19k HAD – ångest
  "VAR36",  # Q19l HAD – depression
  "VAR37",  # Q19m HAD – ångest
  "VAR38",  # Q19n HAD – depression
  "VAR41",  # Q23  ISI – nöjd sömnmönster
  "VAR42",  # Q24  ISI – funktionsförmåga
  "VAR43",  # Q25  ISI – livskvalitet
  "VAR44",  # Q26  ISI – bekymrad sömn
  "VAR45",  # Q27  tid fysisk träning/vecka (6 alternativ)
  "VAR46",  # Q28  dagar fysisk träning (0–7)
  "VAR47",  # Q29  tid vardagsmotion/vecka (7 alternativ)
  "VAR48",  # Q30  dagar vardagsmotion (0–7)
  "VAR49",  # Q31a rökning (4 alternativ)
  "VAR50",  # Q31b antal cigaretter/dag (3 alternativ)
  "VAR51",  # Q32a snus (4 alternativ)
  "VAR52",  # Q32b antal dosor/vecka (3 alternativ)
  "VAR53"   # Q33  kontakt framtida studier: Ja / Nej
)

non_exclusive <- c(
  "VAR12",  # Q7b  smärtlokalisationer (45 kroppsregioner, flera möjliga)
  "VAR20"   # Q14  behandlingsåtgärder (flera möjliga)
)

matrix_groups <- c(
  "VAR10",  # Q6   arbetslivssituation (25/50/75/100 % per rad)
  "VAR19",  # Q13  smärtläkemedel (fritext per rad)
  "VAR21",  # Q15  PSQ smärtkänslighet (0–10 per item)
  "VAR22",  # Q16  CSI del A (1–5 per item)
  "VAR23",  # Q17  CSI del B diagnoser (Nej/Ja per diagnos)
  "VAR24",  # Q18  sjukdomar/skador (Ja/Nej per sjukdom)
  "VAR39",  # Q20–21 SCI-93 (1–5 per item)
  "VAR40"   # Q22  sömnproblem ISI (1–5 per item)
)
```

### Build label table from VariableView

```{r}
label_table <- tibble(
  variable  = vv[["Variable Name"]],
  label     = vv[["Label"]],
  type      = vv[["Type"]],
  data_type = vv[["Data Type"]],
  values    = vv[["Value Codes"]],
  missing   = vv[["Missing Code"]]
) %>%
  mutate(base = sub("_.*", "", variable))
```

### Build group summary

```{r}
group_info <- label_table %>%
  filter(grepl("^VAR[0-9]+_", variable)) %>%
  group_by(base) %>%
  summarise(
    n_items    = n(),
    first_type = first(type),
    question   = sub(" - .*", "", first(label)) %>%
                 str_replace_all("\\r\\n", " ") %>%
                 str_trim(),
    .groups = "drop"
  ) %>%
  mutate(class = case_when(
    base %in% exclusive     ~ "exclusive",
    base %in% non_exclusive ~ "non_exclusive",
    base %in% matrix_groups ~ "matrix",
    TRUE                    ~ "UNCLASSIFIED"
  ))

exclusive_bases     <- group_info$base[group_info$class == "exclusive"]
non_exclusive_bases <- group_info$base[group_info$class == "non_exclusive"]

print(as.data.frame(group_info[, c("base", "n_items", "class")]), row.names = FALSE)
```

### Build codebook

Creates lookup tables mapping each numeric level to its descriptive label.

```{r}
codebook_exclusive <- label_table %>%
  filter(base %in% exclusive_bases) %>%
  mutate(
    sub_index    = as.integer(sub("^VAR[0-9]+_", "", variable)),
    option_label = str_replace_all(label, "\\r\\n", " ") %>% sub("^.* - ", "", .)
  ) %>%
  select(base, sub_index, option_label, variable)

codebook_non_exclusive <- label_table %>%
  filter(base %in% non_exclusive_bases) %>%
  mutate(option_label = str_replace_all(label, "\\r\\n", " ") %>% sub("^.* - ", "", .)) %>%
  select(base, variable, option_label)
```

------------------------------------------------------------------------

## Chapter 3: Verification

Cross-checks the manual classification against VariableView metadata before any data is modified. Produces warnings (not errors) so the script still runs, but issues are printed to the console.

```{r}
verify_classification <- function(group_info) {
  warnings <- character(0)

  uncl <- filter(group_info, class == "UNCLASSIFIED")
  if (nrow(uncl) > 0)
    warnings <- c(warnings, paste0(
      "UNCLASSIFIED groups (add to one of the three vectors): ",
      paste(uncl$base, collapse = ", ")))

  matrix_as_excl <- filter(group_info, class == "exclusive",
                            grepl("Matrix", first_type, ignore.case = TRUE))
  if (nrow(matrix_as_excl) > 0)
    warnings <- c(warnings, paste0(
      "Classified as exclusive but VariableView Type is Matrix: ",
      paste(matrix_as_excl$base, collapse = ", ")))

  excl_multi <- filter(group_info, class == "exclusive",
                       grepl("[Ff]lera alternativ", question))
  if (nrow(excl_multi) > 0)
    warnings <- c(warnings, paste0(
      "Classified as exclusive but label says 'Flera alternativ': ",
      paste(excl_multi$base, collapse = ", ")))

  cb_as_matrix <- filter(group_info, class == "matrix",
                          grepl("Check Boxes", first_type, ignore.case = TRUE))
  if (nrow(cb_as_matrix) > 0)
    warnings <- c(warnings, paste0(
      "Classified as matrix but VariableView Type is Check Boxes: ",
      paste(cb_as_matrix$base, collapse = ", ")))

  many_excl <- filter(group_info, class == "exclusive", n_items > 11)
  if (nrow(many_excl) > 0)
    warnings <- c(warnings, paste0(
      "Exclusive with >11 sub-items (verify single-select): ",
      paste(many_excl$base, " (", many_excl$n_items, ")", sep = "", collapse = ", ")))

  if (length(warnings) == 0) {
    cat("Verification passed: no issues found.\n")
  } else {
    cat("Verification warnings:\n")
    for (w in warnings) cat("  !", w, "\n")
  }

  invisible(warnings)
}

verify_classification(group_info)
```

------------------------------------------------------------------------

## Chapter 4: Data adjustments

Row-level corrections identified during data quality checks. These are applied to the raw data before any recoding.

```{r}
# VAR07: weight 995 kg → 95 kg (data entry error, ID 7)
processed$VAR07[processed$ID == 7] <- 95

# VAR15_8: value 21 → 2 (Nej, data entry error, ID 2005)
processed$VAR15_8[processed$ID == 2005] <- 2

# VAR04: missing legal sex imputed from personnummer.
# Swedish personnummer: the second-to-last digit is odd for men, even for women.
pnr        <- trimws(as.character(processed$VAR02))
sex_digit  <- as.integer(substr(pnr, nchar(pnr) - 1, nchar(pnr) - 1))
is_woman   <- sex_digit %% 2 == 0

sex_missing <- (processed$VAR04_1 != 1 & processed$VAR04_2 != 1) |
               (is.na(processed$VAR04_1) & is.na(processed$VAR04_2))
if (any(sex_missing, na.rm = TRUE)) {
  processed$VAR04_1[sex_missing] <- ifelse(is_woman[sex_missing], 2, 1)
  processed$VAR04_2[sex_missing] <- ifelse(is_woman[sex_missing], 1, 2)
}
```

------------------------------------------------------------------------

## Chapter 5: Standardization of values

Binary exclusive variables are raw-coded **1 = Ja, 2 = Nej**. These are recoded to **1 = yes, 0 = no**. Matrix variables on a **1–5** scale are shifted to **0–4**. Missing codes (999/998) are set to NA.

```{r}
recode_vars <- c(
  "VAR04_", "VAR05_", "VAR08_", "VAR09_",
  paste0("VAR", 11:18, "_"), "VAR20_",
  paste0("VAR", 24:38, "_"),
  paste0("VAR", 41:52, "_")
)

shift_vars <- c("VAR10_", "VAR22_", "VAR23_", "VAR39_", "VAR40_")

processed <- processed %>%
  mutate(
    across(starts_with(recode_vars), ~ replace(.x, .x == 999, NA_real_)),
    across(starts_with(shift_vars),  ~ replace(.x, .x == 999, NA_real_)),
    across(starts_with(recode_vars), ~ replace(.x, .x == 2,   0)),
    across(starts_with(shift_vars),  ~ .x - 1)
  )
```

------------------------------------------------------------------------

## Chapter 6: Formation of new variables

### Age and birthyear

Birthyear is extracted from "personnummer". Swedish personal number can be 10 or 12 digits. Assuming no one is born before 1915, impute "20XX" when numbers are smaller than 15. For 10-digit numbers, YY ≥ 15 → 19YY, else → 20YY.

```{r}
resp_year <- as.integer(format(as.Date(processed$Svarstillfälle), "%Y"))

processed$birthyear <- ifelse(
  nchar(pnr) == 12,
  as.integer(substr(pnr, 1, 4)),
  {
    yy <- as.integer(substr(pnr, 1, 2))
    ifelse(yy >= 15L, 1900L + yy, 2000L + yy)
  }
)

processed$age <- resp_year - processed$birthyear
```

### Sex and gender identity

The variables sex and gender identity will be recoded in Chapter 7 and 8.

### Body Mass Index

BMI categories follow WHO classification (Weir & Jan, *StatPearls*, 2023).

```{r}
processed <- processed %>%
  mutate(
    BMI = if_else(!is.na(VAR06) & VAR06 > 0 & !is.na(VAR07),
                  VAR07 / (VAR06 / 100)^2, NA_real_),

    BMI_4cat = case_when(
      BMI <  18.5             ~ 0,  # underweight
      BMI >= 18.5 & BMI < 25 ~ 1,   # normal
      BMI >= 25   & BMI < 30 ~ 2,   # overweight
      BMI >= 30               ~ 3,  # obese
      TRUE                    ~ NA_real_
    ),

    BMI_6cat = case_when(
      BMI <  18.5             ~ 0,  # underweight
      BMI >= 18.5 & BMI < 25 ~ 1,  # normal
      BMI >= 25   & BMI < 30 ~ 2,  # overweight
      BMI >= 30   & BMI < 35 ~ 3,  # obese I
      BMI >= 35   & BMI < 40 ~ 4,  # obese II
      BMI >= 40               ~ 5,  # obese III
      TRUE                    ~ NA_real_
    )
  )
```

### Education and origin

Education in two categories for Rasch analysis: 0=gymnasium, grundskola and 1=university, högskola. Some other educational variables are recoded in Chapter 7 and 8.

```{r}
# edu_BIN: 0 = non-university, 1 = university
processed$edu_BIN <- processed$VAR08_3

# origin: 0 = Sweden, 1 = other Nordic, 2 = European non-Nordic, 3 = outside Europe
processed$origin <- processed$VAR09_2 + 2 * processed$VAR09_3 + 3 * processed$VAR09_4
```

### Employment and pensioner

Assuming if someone did not fill in any degree of occupation, sick leave or studies, it is equal to 0%. A person is categorized as pensioner if 65 years old and did not fill in any occupation degree.

```{r}
processed <- processed %>%
  mutate(
    VAR10_1 = if_else(is.na(VAR10_1), 0, as.numeric(VAR10_1)),
    VAR10_2 = if_else(is.na(VAR10_2), 0, as.numeric(VAR10_2)),
    VAR10_3 = if_else(is.na(VAR10_3), 0, as.numeric(VAR10_3)),
    pensioner65 = if_else(age >= 65 & VAR10_1 == 0 & VAR10_2 == 0 & VAR10_3 == 0, 1, 0)
  )
```

------------------------------------------------------------------------

### Pain variables

```{r}
processed$pain7days <- processed$VAR11_1  # 1 = yes, 0 = no

processed$pain_duration <- rowSums(processed[, paste0("VAR17_", 1:5)] * 0:4)

processed$chronic_pain <- ifelse(
  processed$pain7days == 1 & processed$pain_duration >= 2, 1, 0)
processed$chronic_pain[is.na(processed$pain7days)] <- NA

processed$persistent <- processed$VAR18_2  # 0 = periodisk, 1 = ihållande

processed$pain_freq <- rowSums(processed[, paste0("VAR16_", 1:5)] * 1:5)

processed$nrs_mean7d <- rowSums(processed[, paste0("VAR13_", 1:11)] * 0:10)
processed$nrs_rest7d <- rowSums(processed[, paste0("VAR14_", 1:11)] * 0:10)
processed$nrs_move7d <- rowSums(processed[, paste0("VAR15_", 1:11)] * 0:10)
```

### Number of pain sites

```{r}
processed$nr_sites <- ifelse(
  processed$pain7days == 1,
  rowSums(processed[, paste0("VAR12_", 1:45)], na.rm = FALSE),
  NA
)
```

### Widespread pain — ACR90

```{r}
processed$q1 <- rowSums(processed[, c("VAR12_5","VAR12_7","VAR12_9","VAR12_11",
                                       "VAR12_26","VAR12_28","VAR12_30","VAR12_32")])
processed$q2 <- rowSums(processed[, c("VAR12_4","VAR12_6","VAR12_8","VAR12_10",
                                       "VAR12_27","VAR12_29","VAR12_31","VAR12_33")])
processed$q3 <- rowSums(processed[, c("VAR12_18","VAR12_20","VAR12_22",
                                       "VAR12_38","VAR12_40","VAR12_42","VAR12_44")])
processed$q4 <- rowSums(processed[, c("VAR12_17","VAR12_19","VAR12_21",
                                       "VAR12_39","VAR12_41","VAR12_43","VAR12_45")])
processed$q5 <- rowSums(processed[, c("VAR12_3","VAR12_12","VAR12_13","VAR12_25",
                                       "VAR12_34","VAR12_35","VAR12_36","VAR12_37")])

processed$contralat_1 <- ifelse(processed$q5 > 0 & processed$q4 > 0 & processed$q1 > 0, 1, 0)
processed$contralat_2 <- ifelse(processed$q5 > 0 & processed$q3 > 0 & processed$q2 > 0, 1, 0)

processed$ACR90         <- ifelse(processed$contralat_1 == 1 | processed$contralat_2 == 1, 1, 0)
processed$ChronicP_ACR90 <- ifelse(processed$chronic_pain == 1 & processed$ACR90 == 1, 1, 0)
```

### Widespread pain — WP2019

```{r}
region1 <- ifelse(processed$VAR12_2==1|processed$VAR12_23==1|processed$VAR12_5==1|
                  processed$VAR12_26==1|processed$VAR12_7==1|processed$VAR12_28==1|
                  processed$VAR12_9==1|processed$VAR12_11==1|processed$VAR12_30==1|
                  processed$VAR12_32==1, 1, 0)
region2 <- ifelse(processed$VAR12_1==1|processed$VAR12_24==1|processed$VAR12_4==1|
                  processed$VAR12_27==1|processed$VAR12_6==1|processed$VAR12_29==1|
                  processed$VAR12_8==1|processed$VAR12_10==1|processed$VAR12_31==1|
                  processed$VAR12_33==1, 1, 0)
region3 <- ifelse(processed$VAR12_18==1|processed$VAR12_38==1|processed$VAR12_40==1|
                  processed$VAR12_20==1|processed$VAR12_22==1|processed$VAR12_42==1|
                  processed$VAR12_44==1, 1, 0)
region4 <- ifelse(processed$VAR12_17==1|processed$VAR12_39==1|processed$VAR12_41==1|
                  processed$VAR12_19==1|processed$VAR12_21==1|processed$VAR12_43==1|
                  processed$VAR12_45==1, 1, 0)
region5 <- ifelse(processed$VAR12_3==1|processed$VAR12_25==1|processed$VAR12_34==1|
                  processed$VAR12_35==1|processed$VAR12_36==1|processed$VAR12_37==1|
                  processed$VAR12_12==1|processed$VAR12_13==1|processed$VAR12_14==1|
                  processed$VAR12_15==1, 1, 0)
nr_regions <- region1 + region2 + region3 + region4 + region5

processed$WPI_AGE <-
  ifelse(processed$VAR12_2==1|processed$VAR12_23==1,1,0) +
  ifelse(processed$VAR12_5==1|processed$VAR12_26==1,1,0) +
  ifelse(processed$VAR12_7==1|processed$VAR12_28==1,1,0) +
  ifelse(processed$VAR12_9==1|processed$VAR12_11==1|
         processed$VAR12_30==1|processed$VAR12_32==1,1,0) +
  ifelse(processed$VAR12_1==1|processed$VAR12_24==1,1,0) +
  ifelse(processed$VAR12_4==1|processed$VAR12_27==1,1,0) +
  ifelse(processed$VAR12_6==1|processed$VAR12_29==1,1,0) +
  ifelse(processed$VAR12_8==1|processed$VAR12_10==1|
         processed$VAR12_31==1|processed$VAR12_33==1,1,0) +
  ifelse(processed$VAR12_18==1|processed$VAR12_38==1,1,0) +
  ifelse(processed$VAR12_18==1|processed$VAR12_40==1,1,0) +
  ifelse(processed$VAR12_20==1|processed$VAR12_22==1|
         processed$VAR12_42==1|processed$VAR12_44==1,1,0) +
  ifelse(processed$VAR12_17==1|processed$VAR12_39==1,1,0) +
  ifelse(processed$VAR12_17==1|processed$VAR12_41==1,1,0) +
  ifelse(processed$VAR12_19==1|processed$VAR12_21==1|
         processed$VAR12_43==1|processed$VAR12_45==1,1,0) +
  ifelse(processed$VAR12_3==1|processed$VAR12_25==1,1,0) +
  ifelse(processed$VAR12_34==1|processed$VAR12_35==1,1,0) +
  ifelse(processed$VAR12_36==1|processed$VAR12_37==1,1,0) +
  ifelse(processed$VAR12_12==1|processed$VAR12_13==1,1,0) +
  ifelse(processed$VAR12_14==1|processed$VAR12_15==1,1,0)

processed$WPI_AGE <- ifelse(is.na(processed$pain7days), NA, processed$WPI_AGE)
processed$WPI_AGE <- ifelse(is.na(processed$WPI_AGE) & processed$pain7days == 0, 0, processed$WPI_AGE)

processed$WP2019_AGE    <- as.integer(processed$WPI_AGE > 6 & nr_regions > 3)
processed$CP_WP2019_AGE <- as.integer(processed$chronic_pain == 1 & processed$WP2019_AGE == 1)

processed$Pain3cat_CWSP2019 <- ifelse(
  processed$chronic_pain == 1 & processed$CP_WP2019_AGE == 1, 2,
  ifelse(processed$chronic_pain == 1, 1, 0))

rm(region1, region2, region3, region4, region5, nr_regions)
```

### Treatments (VAR20)

VAR20 is non-exclusive (multiple treatments possible). NA is treated as 0 (no treatment reported).

```{r}
for (col in paste0("VAR20_", c(1,2,3,5,6))) {
  processed[[col]][is.na(processed[[col]])] <- 0
}

processed$trt_mmr       <- processed$VAR20_1
processed$trt_neurostim  <- processed$VAR20_2
processed$trt_surgery    <- processed$VAR20_3
processed$trt_opioid     <- processed$VAR20_5
processed$trt_other      <- processed$VAR20_6
```

### PSQ (Pain Sensitivity Questionnaire)

PSQ-14 uses items 1–4, 6–8, 10–12, 14–17 (excludes items 5, 9, 13). PSQ-17 uses all 17 items.

```{r}
processed$psq14 <- rowMeans(processed[, paste0("VAR21_", c(1:4, 6:8, 10:12, 14:17))],
                             na.rm = FALSE)

processed$psq17 <- rowMeans(processed[, paste0("VAR21_", 1:17)],
                             na.rm = FALSE)
```

### CSI (Central Sensitization Inventory)

CSI part A: 25 items, each shifted to 0–4. Sum range 0–100.

```{r}
processed$csi <- rowSums(processed[, paste0("VAR22_", 1:25)])
```

CSI part B: physician diagnoses, shifted to 0 (No) / 1 (Yes).

```{r}
csi_b_names <- c("restlesslegs", "chronic_fatigue", "fibromyalgia", "TMD",
                  "headache_migraine", "ibs", "chemical_sens", "neckpain",
                  "anxiety", "depression")

for (i in seq_along(csi_b_names)) {
  processed[[csi_b_names[i]]] <- processed[[paste0("VAR23_", i)]]
}
```

------------------------------------------------------------------------

### HAD (Hospital Anxiety and Depression Scale)

The HADS consists of 14 items: 7 for anxiety (odd questions, VAR25–VAR37) and 7 for depression (even questions, VAR26–VAR38). Each item is scored 0–3; subscale range 0–21. Higher scores indicate greater distress.

**Scoring of positive vs. negative items**

The HADS contains both negative items (agreeing most indicated the individual is much depressed or have much anxiety) where the most distressing response is the first option, and positive items (agreeing most is an indication of low degree of depression or anxiety) where the healthiest response is the first option. To ensure that all items contribute in the same direction (higher = more distress), positive items are reverse-scored.

*Sources:*

- Zigmond AS, Snaith RP. The hospital anxiety and depression scale. *Acta Psychiatrica Scandinavica*, 1983;67(6):361–370.

**Item classification:**

| VAR | Item | Subscale | Type | \_1 option | \_1 weight |
|------------|------------|------------|------------|------------|------------|
| VAR25 | Jag känner mig spänd och nervös | Anxiety | Negative | Mestadels | 3 |
| VAR26 | Jag uppskattar fortfarande saker... | Depression | Positive | Definitivt lika mycket | 0 |
| VAR27 | Jag har en känsla av att något hemskt... | Anxiety | Negative | Mycket klart och obehagligt | 3 |
| VAR28 | Jag kan skratta och se det roliga... | Depression | Positive | Lika ofta som tidigare | 0 |
| VAR29 | Jag bekymrar mig över saker | Anxiety | Negative | Mestadels | 3 |
| VAR30 | Jag känner mig på gott humör | Depression | Positive | Mestadels | 0 |
| VAR31 | Jag kan sitta stilla och känna mig avslappnad | Anxiety | Positive | Absolut | 0 |
| VAR32 | Allting känns trögt | Depression | Negative | Nästan alltid | 3 |
| VAR33 | Jag känner mig orolig, som om jag hade fjärilar... | Anxiety | Negative | Väldigt ofta | 3 |
| VAR34 | Jag har tappat intresset för hur jag ser ut | Depression | Negative | Fullständigt | 3 |
| VAR35 | Jag känner mig rastlös | Anxiety | Negative | Väldigt ofta | 3 |
| VAR36 | Jag ser med glädje fram emot saker och ting | Depression | Positive | Lika mycket som tidigare | 0 |
| VAR37 | Jag får plötsliga panikkänslor | Anxiety | Negative | Väldigt ofta | 3 |
| VAR38 | Jag kan uppskatta en god bok, ett TV-program... | Depression | Positive | Ofta | 0 |

```{r}
# N-items: _1 = most distress → weight 3
# After recoding (2→0), _k = 1 if selected, 0 otherwise.
score_had_N <- function(var) {
  3 * processed[[paste0(var, "_1")]] +
  2 * processed[[paste0(var, "_2")]] +
  1 * processed[[paste0(var, "_3")]] +
  0 * processed[[paste0(var, "_4")]]
}

# P-items: _1 = healthiest response → weight 0 (reverse-scored)
score_had_P <- function(var) {
  0 * processed[[paste0(var, "_1")]] +
  1 * processed[[paste0(var, "_2")]] +
  2 * processed[[paste0(var, "_3")]] +
  3 * processed[[paste0(var, "_4")]]
}

# had_a: anxiety subscale (0–21)
# Items: VAR25(N), VAR27(N), VAR29(N), VAR31(P), VAR33(N), VAR35(N), VAR37(N)
processed$had_a <-
  score_had_N("VAR25") +   # a) N – spänd och nervös
  score_had_N("VAR27") +   # c) N – känsla av att något hemskt händer
  score_had_N("VAR29") +   # e) N – bekymrar sig
  score_had_P("VAR31") +   # g) P – kan sitta stilla och slappna av
  score_had_N("VAR33") +   # i) N – fjärilar i magen
  score_had_N("VAR35") +   # k) N – rastlös
  score_had_N("VAR37")     # m) N – panikkänslor

# had_d: depression subscale (0–21)
# Depression items: VAR26(P), VAR28(P), VAR30(P), VAR32(N), VAR34(N), VAR36(P), VAR38(P)
processed$had_d <-
  score_had_P("VAR26") +   # b) P – uppskattar fortfarande saker
  score_had_P("VAR28") +   # d) P – kan skratta och se det roliga
  score_had_P("VAR30") +   # f) P – på gott humör
  score_had_N("VAR32") +   # h) N – allting känns tungt
  score_had_N("VAR34") +   # j) N – tappat intresset för utseende
  score_had_P("VAR36") +   # l) P – ser med glädje framemot saker
  score_had_P("VAR38")     # n) P – kan uppskatta bok/TV/radio

summary(processed$had_a)
summary(processed$had_d)
```

------------------------------------------------------------------------

### SCI-93

35 items, each shifted from 1–5 to 0–4. Sum range 0–140.

```{r}
processed$sci93 <- rowSums(processed[, paste0("VAR39_", 1:35)])
```

### ISI (Insomnia Severity Index)

VAR40_1–3: sleep items shifted to 0–4 (via shift_vars above). Sum range 0–12. VAR41–44: 5-option Likert items recoded to 0/1; scored 0–4 each. Total range 0–28.

```{r}
processed$ISI <-
  rowSums(processed[, paste0("VAR40_", 1:3)]) +
  processed$VAR41_1*0 + processed$VAR41_2*1 + processed$VAR41_3*2 +
    processed$VAR41_4*3 + processed$VAR41_5*4 +
  processed$VAR42_1*0 + processed$VAR42_2*1 + processed$VAR42_3*2 +
    processed$VAR42_4*3 + processed$VAR42_5*4 +
  processed$VAR43_1*0 + processed$VAR43_2*1 + processed$VAR43_3*2 +
    processed$VAR43_4*3 + processed$VAR43_5*4 +
  processed$VAR44_1*0 + processed$VAR44_2*1 + processed$VAR44_3*2 +
    processed$VAR44_4*3 + processed$VAR44_5*4
```

------------------------------------------------------------------------

### Lifestyle

#### Physical activity

```{r}
processed$PE_time <- rowSums(processed[, paste0("VAR45_", 1:6)] * 0:5)
processed$PE_days <- rowSums(processed[, paste0("VAR46_", 1:8)] * 0:7)
processed$PA_time <- rowSums(processed[, paste0("VAR47_", 1:7)] * 0:6)
processed$PA_days <- rowSums(processed[, paste0("VAR48_", 1:8)] * 0:7)
```

#### Smoking

```{r}
processed$smoke     <- rowSums(processed[, paste0("VAR49_", 1:4)] * 0:3)
processed$ant_smoke <- rowSums(processed[, paste0("VAR50_", 1:3)] * 1:3)
processed$ant_smoke[processed$smoke < 3] <- 0
```

#### Snuff

```{r}
processed$snuff     <- rowSums(processed[, paste0("VAR51_", 1:4)] * 0:3)
processed$ant_snuff <- rowSums(processed[, paste0("VAR52_", 1:3)] * 1:3)
processed$ant_snuff[processed$snuff < 3] <- 0
```

------------------------------------------------------------------------

## Chapter 7: Collapsing exclusive dummies

Collapses mutually exclusive dummy columns (e.g., VAR04_1, VAR04_2) into a single 0-based categorical column (e.g., VAR04 = 0 or 1). Conflicts, where more than one option is marked, are flagged as **-1** rather than silently dropped as NA, making data errors visible and traceable.

```{r}
collapse_dummies <- function(df, prefix) {
  cols <- grep(paste0("^", prefix, "_"), names(df), value = TRUE)
  cols <- cols[order(as.numeric(sub(paste0(prefix, "_"), "", cols)))]

  result <- apply(df[cols], 1, function(row) {
    hit <- which(row == 1)
    if (length(hit) == 1) return(hit - 1)   # 0-based
    if (length(hit) == 0) return(NA)         # unanswered
    return(-1)                               # conflict: multiple options marked
  })

  df[[cols[1]]] <- result
  names(df)[names(df) == cols[1]] <- prefix
  df[cols[-1]] <- NULL
  df
}
```

HAD variables (VAR25–38) are scored above and then collapsed here. The collapsed value (0–3) represents the selected response option and can be used for descriptive purposes; the scored subscales `had_a` and `had_d` are used for analysis.

```{r}
variables_to_collapse <- c(
  "VAR04", "VAR05",
  "VAR08", "VAR09",
  "VAR11", "VAR13", "VAR14", "VAR15",
  "VAR16", "VAR17", "VAR18",
  "VAR25", "VAR26", "VAR27", "VAR28", "VAR29", "VAR30", "VAR31",
  "VAR32", "VAR33", "VAR34", "VAR35", "VAR36", "VAR37", "VAR38",
  "VAR41", "VAR42", "VAR43", "VAR44",
  "VAR45", "VAR46", "VAR47", "VAR48",
  "VAR49", "VAR50", "VAR51", "VAR52",
  "VAR53"
)

for (v in variables_to_collapse) {
  processed <- collapse_dummies(processed, v)
}
```

### Data verification — conflict check

After collapsing, any row with value -1 indicates a respondent who marked multiple options in a single-select question. These are printed below for manual review.

```{r}
conflict_cols <- intersect(variables_to_collapse, names(processed))
conflict_counts <- processed %>%
  select(all_of(conflict_cols)) %>%
  summarise(across(everything(), ~ sum(. == -1, na.rm = TRUE)))

any_conflicts <- any(conflict_counts > 0)
if (any_conflicts) {
  cat("DATA WARNING: rows with multiple selections in exclusive groups:\n")
  for (col in names(conflict_counts)[conflict_counts > 0]) {
    cat(sprintf("  %s: %d conflicting rows\n", col, conflict_counts[[col]]))
  }
} else {
  cat("Data verification passed: no conflicts found.\n")
}
```

------------------------------------------------------------------------

## Chapter 8: Rename variables and apply labels

```{r}
final <- processed %>%
  rename(
    consent                    = VAR00,
    name                       = VAR01,
    personNumber               = VAR02,
    mailAdress                 = VAR03,
    sexLegal                  = VAR04,
    gender                     = VAR05,
    height                     = VAR06,
    weight                     = VAR07,
    education                  = VAR08,
    birthCountry               = VAR09,
    timeOnWork                 = VAR10_1,
    timeOnSickleave            = VAR10_2,
    timeOnStudies              = VAR10_3,
    painOrAcheLast7Days        = VAR11,
    painIntensityMean7Days     = VAR13,
    painIntensityRest7Days     = VAR14,
    painIntensityMovement7Days = VAR15,
    painFreq12Months           = VAR16,
    firstPainDuration          = VAR17,
    painPersistence            = VAR18,
    painTreatment12Months      = VAR20,
    CSIDiagnosedByDoctor = VAR23,
    healthRelatedIsDiagnosed = VAR24,
    HAD1TenseOrWoundUp = VAR25,
    HAD2EnjoysThingsIUsedTo = VAR26,
    HAD3FrightFeelingSomethingBadWillHappen = VAR27,
    HAD4LaughAndSeeFunnySideInThings = VAR28,
    HAD5WorryingThoughtsThroughMinf = VAR29,
    HAD6IsCheerful = VAR30,
    HAD7AtEaseAndRelax = VAR31,
    HAD8FeelsSlowedDown = VAR32,
    HAD9FeelsButterflyStomach = VAR33,
    HAD10NoInterestInAppearance = VAR34,
    HAD11RestlessNeedToMove = VAR35,
    HAD12LookForwardEnjoyment = VAR36,
    HAD13SuddenPanicFeeling = VAR37,
    HAD14EnjoyGoodBookOrRadioOrTv = VAR38,
    sleep22TroubleFallingasleep = VAR40_1,
    sleep22TroubleStayingAsleep = VAR40_2,
    sleep22WakingUppTooEarly = VAR40_3,
    sleepScheduleSatisfaction = VAR41,
    badSleepAffectsDayFunction = VAR42,
    noticeBadSleepLifeQuality = VAR43,
    worriedAboutSleepProblem = VAR44,
    physActHowMuchTimeShortBreath = VAR45,
    physActAtLeast30MinShortBreath = VAR46,
    physActTimeOnEverydayExercise = VAR47,
    physActDaysSpent30MinExercise = VAR48,
    smoker = VAR49,
    howManyCigsPerDay = VAR50,
    snuser = VAR51,
    howManySnusPacksPerWeek = VAR52,
    canWeContactLater = VAR53,
      
      
  ) %>%
  rename_with(
    ~ paste0("markedBody", str_extract(.x, "\\d+$")),
    matches("^VAR12_\\d+$")
  ) 

# TO BE HELPED
# VAR12, VAR19, VAR21, VAR22, VAR39
```

Apply variable labels from VariableView to remaining columns.

```{r}
lookup <- setNames(
  str_replace_all(label_table$label, "\\r\\n", " "),
  label_table$variable
)
matched <- intersect(names(final), names(lookup))
var_label(final[matched]) <- lookup[matched]

for (b in exclusive_bases) {
  if (b %in% names(final)) {
    q <- group_info$question[group_info$base == b]
    if (length(q) == 1) var_label(final[[b]]) <- q
  }
}
```

------------------------------------------------------------------------

## Chapter 9: Export

The processed dataset is exported to `docs/PainSense_processed_data.xlsx` with two sheets: **Data** (one row per respondent) and **Codebook** (variable labels and response options). The `docs/` folder is excluded from version control since it may contain personal data.

```{r}
dir.create("docs", showWarnings = FALSE)

wb <- createWorkbook()

addWorksheet(wb, "Data")
writeData(wb, "Data", final)

addWorksheet(wb, "Codebook")
writeData(wb, "Codebook", codebook_exclusive)

saveWorkbook(wb, "docs/PainSense_processed_data.xlsx", overwrite = TRUE)

cat(sprintf("Exported %d rows and %d variables to docs/PainSense_processed_data.xlsx\n",
            nrow(final), ncol(final)))
```